The Talking Heads Experiment

### 2015-04-10 chap4/chap4.tex
cumulative changes to preface and chapter 1
updated labels, image widths, quotation marks, capitalisation, etc. for chapters 1-7

%\documentclass{../_combined/fcg-book}
\chapter{The Discrimination Game}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/tree}}
\caption{\label{trees} A discrimination tree contains
a set of categorisers which categorise by checking whether
a sensory value falls in the region of one category or 
not. The discrimination tree shown operates on
values on the GRAY channel.}
\end{figure}

As mentioned earlier, I will label categories
by using the sensory channel from which a category operates, 
followed by the minimum and maximum values of the region
carved out by the category. For example, 
[AREA 0.0--0.25] carves out the region [0.0,0.25] of the 
area channel. When it must be emphasised that 
a category belongs to a particular agent, for 
example {\bfshape a1}, I will write [AREA 0.0--0.25]$_{\mathbf a1}$.

Conjunctive combinations of categories also have
dedicated categorisers which are linked to 
the categorisers of their components (see 
\figref{disnet}). A conjunctive combination 
often yields a more efficient way to pick out 
the topic compared to a single, possibly very fine-grained
distinction. For example, it might be that 
[AREA 0.0--0.25] {\itshape and} [GRAY 0.5-1.0]
together are distinctive but none of the two on their
own is. Other logical combinations are equally of interest
but I will restrict my attention to conjunctive 
combinations. Conjunctive combinations of categories 
will be written between curly brackets (\{,\}) as in 
\{[AREA 0.0--0.25] [GRAY 0.5-1.0]\}.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/disnet}}
\caption{\label{disnet} More complex categorisers
(shown as circles) are formed from the combination
of primitive categorisers.}
\end{figure}

@@

\is{Discrimination Game}Here is a game, which I call the Discrimination Game, which is useful 
to study categorisation in 
a systematic way.\footnote{
The Discrimination Game model together with the 
discrimination trees and its growth dynamics 
was presented for the first time in \cite{Steels:1996}. 
Based on this paper a new implementation of
single category discrimination was implemented by Angus McIntyre 
within the BABEL environment. Later on, Joris Van Looveren
re-implemented the use of conjunctive combinations.}
The game is played by a single agent, randomly 
drawn from a population of agents, and 
is equivalent to the conceptualisation phase of 
the guessing game. The agent perceives the scene and 
chooses a topic from the possible objects segmented in the
scene. He then uses his discrimination trees, as developed so
far, to come up with a category or a conjunction
of categories that is valid for the topic, but not for any
other object in the context. The Discrimination Game succeeds if 
the agent has found distinctive categories, otherwise the game
fails. When the game succeeds, the success counter
of the categorisers involved go up.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game5}}
\caption{\footnotesize \label{geom} A computer generated scene 
from the GEOM world.}
\end{figure}

@@

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [GRAY-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [VPOS 0.0--05]
(lower), and [HPOS 0.0--0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

Imagine that for each object in
the scene and for each sensory channel, there is a ball
containing the value for that object on that channel. It
is introduced in the top categoriser of the 
discrimination tree associated with that channel. 
For example, suppose that there are three
objects in the scene: $O1$, $O2$, and $O3$, 
with gray-scale values 0.6, 0.4, and 0.9 respectively. 
We can therefore imagine three balls labeled with these datavalues
which are input to the top categoriser of the 
GRAY discrimination tree (\figref{balls}). 
A categoriser divides the balls in two bins, 
those that fall in the range of 
one category and those that fall into the range of the other
category. In this case, the left bin contains \{$O2$\} 
(category [GRAY 0.0--0.5]) 
and the right bin \{$O1$, $O3$\} (category [GRAY 0.5-1.0]).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/balls}}
\caption{\footnotesize \label{balls} Balls containing
the data value for the different objects
$O1$, $O2$, $O3$ in the scene. $O1$ is the topic.
$O1$ has the value 0.6 on the GRAY-channel, $O2$ has
0.4 and $O3$ 0.9.} 
\end{figure}

A distinctive category is found when the ball of the 
topic is the only one left in one of the bins. 
If the topic is $O1$, then this is not yet the 
case, because it is together with $O3$ in the 
bin of [GRAY 0.5-1.0]. However when the next categoriser
is exercised, it splits the set \{$O1$, $O3$\} 
into two subsets: \{$O1$\} for [GRAY 0.5--0.75]
and \{$O3$\} for [GRAY-0.75,1.0]. 
[GRAY 0.5--0.75] is a discriminating category because only 
$O1$ is left in its bin.

@@

A realistic agent has hundreds of sensory
channels, and humans probably have tens of thousands
of them, probably grouped with respect to the domains
to which they apply.\footnote{
This is strongly suggested by psychological 
data on the presence of conceptual spaces in human 
categorisation. See \cite{Gardenfors:1999}.} There are discrimination trees
for each of 
these channels or for combinations of them, and 
each can possibly yield a distinctive
category. The categorisation process can therefore be
envisioned like a huge Pachinko hall, in which
balls are trickling down in parallel in hundreds or 
thousands of machines. It is highly 
likely that more than one solution is found
when there are a lot of trees, particularly 
if combinations of categories are allowed as well. 
So, an additional competitive process must 
take place to rank categories even 
though multiple solutions are offered to subsequent
verbalisation processes. We see therefore the same
characteristics as for the perceptual layer, and thus
the `sieve architecture' also 
applies for the conceptual layer (\figref{f:sieve}).

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (square) & 0.413 & 0.317  \\ 
1 (circle) & 0.410 & 0.410 \\ 
2 (square) & 0.163 & 0.163 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game8} Sensory data for the scene in \figref{scene2}.}
\end{center}
\end{table}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene8}}
\caption{\label{scene2} Top: The scene used in Game 8. 
Shape 0 is the topic. Bottom: The discrimination trees available
for this game.}
\end{figure}
The game is reported by the commentator as follows: 
\begin{verbatim}
Game 8
 a1 segments the context in 3 objects: 
    square-0, circle-1, square-2
 a1 chooses square-0 as the topic
 The discrimination game fails
\end{verbatim}
The game fails because for the two sensory channels 
for which there are discrimination trees, the values 
of the segments are all within the lower range and so 
no distinctive category or category set could be found. 
This failure stimulates
the discrimination network to expand, but any node, 
including a top node of some of the other sensory channels 
can be chosen for further expansion.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene22}}
\caption{\label{scene3} Top: The scene used in Game 22. 
Bottom: The discrimination trees available to the
agent.}
\end{figure}
The next example shows a game (Game 22) based on the scene 
in \figref{scene3}. The topic is the triangle, shape 0. 
The discrimination trees for 
HEIGHT and WIDTH have already more than one level and 
a discrimination tree for HPOS has developed.
The relevant sensor-scaled data for these three sensory
channels is shown in \tabref{tab:t-game22}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.167 & 0.437 & 0.573  \\ 
1 (rectangle) & 0.789 & 0.563 & 0.287 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22} Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}

The scene is very simple so there are several possible
solutions: The triangle is more to the left, it is less wide 
and taller. Each of these possibilities is discovered
and their score (purely based on past performance) is 
looked up. The game is reported by the commentator
as follows: 
\begin{verbatim}
Game 22
 a1 segments the scene in 2 objects:  
   triangle-0, rectangle-1
 a1 chooses triangle-0 as the topic 
 a1 categorises the topic as [HPOS 0.0--0.5] (score 0.57), 
   [HEIGHT 0.5-1.0] (score 0.09), or 
   [WIDTH 0.0--0.5] (score 0.0) 
 The discrimination game succeeds
\end{verbatim}

@@

Game 22 shows at once why context-scaling 
and saliency is important. When we inspect the scene in 
\figref{scene3}, we do not quite see so clearly 
that the triangle is less wide than the square, so why
is [WIDTH 0.0--0.5] 
nevertheless considered? Examination of the data shows
that the WIDTH values, 0.437 for the triangle
and 0.563 for the square, are very close 
to each other, but just by luck fall within 
the two regions carved out by the WIDTH discrimination
tree. On the other hand, the values on the HPOS channel 
are much further apart and so they are preferred.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
 0.622 & 0.125 & 0.286  \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22-sal} Sensory data for the scene in Game 22.}
\end{center}
\end{table}
From this table we see immediately that HPOS is the 
most salient channel and should be preferred by far, followed
by HEIGHT and then WIDTH. Thus we can expect the
agent to choose HPOS based on saliency. When the saliency threshold
is set to a reasonably high value, the other channels
would not even be considered, they would not pass the 
sieve of the perceptual layer.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.0 & 0.0 & 1.0  \\ 
1 (rectangle) & 1.0 & 1.0 & 0.0 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22scaled} Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}
So the best thing to do (and this is what the Talking Heads 
effectively do) is to first perform sensor-scaling, then 
compute saliency to determine which channel should be 
preferred, then perform context-scaling, to get clearly 
distinguished sensory values, and then do categorisation. 
Note that context-scaling has the same effect as using 
prototype-based categorisation because the actual values 
are pulled towards extremes, and thus perceived as 
prototypes. Context-scaling is not always desirable. For example, 
in the case of colour categorisation the actual channel
data should be maintained because here categorisation 
takes place on the basis of actual values.

@@

The next game (Game 24) is based on the scene in 
\figref{scene4}. The topic is triangle-0. 
The discrimination trees are the same as for
Game 22. The game (based on sensor-scaled values)
succeeds with a conjunctive combination of two
categories: 
\begin{verbatim}
Game 24
 a1 segments the scene in 4 objects: 
   triangle-0, triangle-1, square-2, rectangle-3 
 a1 chooses triangle-0 as topic
 a1 categorises the topic as 
   {[HEIGHT 0.0--0.5] [WIDTH 0.5-1.0]}
 The discrimination game succeeds
\end{verbatim}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game24}}
\caption{\label{scene4} The scene used in Game 24.}
\end{figure}
The relevant data after sensor-scaling for the 
two sensory channels involved are shown in \tabref{tab:t-game24}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (triangle) & 0.170 & 0.513  \\ 
1 (triangle) & 0.653 & 0.570 \\  
2 (square) & 0.213 & 0.213 \\  
3 (rectangle) & 0.613 & 0.310 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game24} Sensory data for the scene in \figref{scene4}.}
\end{center}
\end{table}

[HEIGHT 0.0--0.5] is valid for triangle-0 and square-2 but
filters out the other segments. 
[WIDTH 0.5-1.0] is valid for triangle-0 and triangle-1 and 
filters out the others.  
The conjunctive combination of these two categories only 
retains triangle-0 and is therefore the one that is chosen.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & Saliency\\ \midrule
HPOS & 0.27 & 0.16 & 0.11\\ 
VPOS & 0.20 & 0.20 & 0.0\\ 
HEIGHT & 0.15 & 0.15 & 0.0\\ 
WIDTH & 0.10 & 0.11 & 0.01\\ 
AREA & 0.10 & 0.10 & 0.0\\ 
R & 0.23 & 0.25 & 0.02\\ 
G & 0.32 & 0.34  & 0.02\\ 
B & 0.63 & 0.65 & 0.02\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-real} Sensory data from a real world scene with segmentation shown in \ref{f:plate-10}.}
\end{center}
\end{table}
Clearly HPOS is the most salient channel and should be 
prefered by the agent. When performing context scaling, 
the two values for HPOS are drawn apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [HPOS 0.5-1.0] (to the 
right) easily distinguishes the topic (object-0) from
object-1. The game is reported as follows: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5-1.0]
\end{verbatim}
This example illustrates well why the categorisation 
of the Talking Heads is so robust and why agents 
often share the same conceptualisation even if the details  
of their raw perception is quite different. 
The saliency factor helps to focus the agents on those
aspects of the scene that stand out. There is 
an enormous reduction of variation, first by scaling then by 
the categorisation process itself.

@@

A new subcategoriser is constructed by taking a
categoriser node in the tree and dividing its range into 
two new subranges and thus two new subcategorisers.
For example, if there is a 
categoriser [HPOS 0.0--0.5], which triggers when the 
object is in the left most half of a scene, i.e. with HPOS
within $[0.0,0.5]$, then two 
subcategories are created by dividing $[0.0,0.5]$
into two halves, one for the 
range $[0.0,0.25]$ ([HPOS 0.0--0.25] or totally left) 
and one for the range $[0.25,0.5]$ ([HPOS 0.25--0.5] or mid-left). 
A new categoriser is added to the tree for each of these
halves.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap4/figs/avsuc}}
\caption{\label{avsuc} The graph displays the 
average success per 25 Discrimination Games for a
series of 500 games played by a single agent. Success climbs
to 100\%. The graph also displays the size of the 
agent's repertoire of categories. Each scene contains
between 3 and 6 objects.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.75\textwidth]{chap4/figs/dynatrees}}
\caption{\label{dyna-trees} Some snapshots of an evolving 
repertoire in a single agent using growth and pruning. 
Trees are shown after 500 games (a), 1000 games (b), 1500 games
(c) and 2000 games (d).} 
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/adptwrl1}}
\caption{\label{adptwrl1} Two snapshots 
from a series of 500 discrimination 
games developing categories to distinguish rectangles of 
the same average graylevel. Channel 2 (the ratio channel) 
and channel 3 (the grayscale channel) do not develop.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/adptwrl2}}
\caption{\label{adptwrl2} Two snapshots from 
an additional series of 500 Discrimination Games after
the environment has become more complex. RATIO
(channel 2) and GRAY (channel 3) have started to develop.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/adptwrlglo}}
\caption{\label{adptwrlglo} Evolution of success and ontology
size in a series of 1000 Discrimination Games played by a single agent.
In phase 1, only rectangles of the same graylevel are 
generated by the GEOM world. In phase 2, additional types of 
shapes are generated by the environment.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/discri200}}
\caption{\label{discri200a} The discrimination trees 
developed by two physically embodied agents 
{\bfshape a1} (left) and {\bfshape a2} (right). The top of the 
figure shows the trees after 
playing 100 games and bottom after 200 games.}
\end{figure}

>>>>

%\documentclass{../_combined/fcg-book}
\chapter{The Discrimination Game}
\label{chap:4}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/tree.pdf}}
\caption{\label{trees}A discrimination tree contains
a set of categorisers which categorise by checking whether
a sensory value falls in the region of one category or 
not. The discrimination tree shown operates on
values on the GRAY channel.}
\end{figure}

As mentioned earlier, I will label categories
by using the sensory channel from which a category operates, 
followed by the minimum and maximum values of the region
carved out by the category. For example, 
[AREA 0.0–0.25] carves out the region [0.0,0.25] of the 
area channel. When it must be emphasised that 
a category belongs to a particular agent, for 
example {\bfshape a1}, I will write [AREA 0.0–0.25]$_{\mathbf a1}$.

Conjunctive combinations of categories also have
dedicated categorisers which are linked to 
the categorisers of their components (see 
\figref{disnet}). A conjunctive combination 
often yields a more efficient way to pick out 
the topic compared to a single, possibly very fine-grained
distinction. For example, it might be that 
[AREA 0.0–0.25] {\itshape and} [GRAY 0.5–1.0]
together are distinctive but none of the two on their
own is. Other logical combinations are equally of interest
but I will restrict my attention to conjunctive 
combinations. Conjunctive combinations of categories 
will be written between curly brackets (\{,\}) as in 
\{[AREA 0.0–0.25] [GRAY 0.5–1.0]\}.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/disnet.pdf}}
\caption{\label{disnet}More complex categorisers
(shown as circles) are formed from the combination
of primitive categorisers.}
\end{figure}

@@

\is{Discrimination Game}Here is a game, which I call the Discrimination Game, which is useful 
to study categorisation in 
a systematic way.\footnote{
The Discrimination Game model together with the 
discrimination trees and its growth dynamics 
was presented for the first time in \cite{Steels:1996}. 
Based on this paper a new implementation of
single category discrimination was implemented by Angus McIntyre 
within the \textsc{babel} environment. Later on, Joris Van Looveren
re-implemented the use of conjunctive combinations.}
The game is played by a single agent, randomly 
drawn from a population of agents, and 
is equivalent to the conceptualisation phase of 
the guessing game. The agent perceives the scene and 
chooses a topic from the possible objects segmented in the
scene. He then uses his discrimination trees, as developed so
far, to come up with a category or a conjunction
of categories that is valid for the topic, but not for any
other object in the context. The Discrimination Game succeeds if 
the agent has found distinctive categories, otherwise the game
fails. When the game succeeds, the success counter
of the categorisers involved go up.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game5.pdf}}
\caption{\label{geom}A computer generated scene 
from the GEOM world.}
\end{figure}

@@

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [GRAY-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [VPOS 0.0--05]
(lower), and [HPOS 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

Imagine that for each object in
the scene and for each sensory channel, there is a ball
containing the value for that object on that channel. It
is introduced in the top categoriser of the 
discrimination tree associated with that channel. 
For example, suppose that there are three
objects in the scene: $O1$, $O2$, and $O3$, 
with gray-scale values 0.6, 0.4, and 0.9 respectively. 
We can therefore imagine three balls labeled with these datavalues
which are input to the top categoriser of the 
GRAY discrimination tree (\figref{balls}). 
A categoriser divides the balls in two bins, 
those that fall in the range of 
one category and those that fall into the range of the other
category. In this case, the left bin contains \{$O2$\} 
(category [GRAY 0.0–0.5]) 
and the right bin \{$O1$, $O3$\} (category [GRAY 0.5–1.0]).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/balls.pdf}}
\caption{\label{balls}Balls containing
the data value for the different objects
$O1$, $O2$, $O3$ in the scene. $O1$ is the topic.
$O1$ has the value 0.6 on the GRAY-channel, $O2$ has
0.4 and $O3$ 0.9.} 
\end{figure}

A distinctive category is found when the ball of the 
topic is the only one left in one of the bins. 
If the topic is $O1$, then this is not yet the 
case, because it is together with $O3$ in the 
bin of [GRAY 0.5–1.0]. However when the next categoriser
is exercised, it splits the set \{$O1$, $O3$\} 
into two subsets: \{$O1$\} for [GRAY 0.5–0.75]
and \{$O3$\} for [GRAY-0.75,1.0]. 
[GRAY 0.5–0.75] is a discriminating category because only 
$O1$ is left in its bin.

@@

A realistic agent has hundreds of sensory
channels, and humans probably have tens of thousands
of them, probably grouped with respect to the domains
to which they apply.\footnote{
This is strongly suggested by psychological 
data on the presence of conceptual spaces in human 
categorisation. See \cite{Gardenfors:1999}.} There are discrimination trees
for each of 
these channels or for combinations of them, and 
each can possibly yield a distinctive
category. The categorisation process can therefore be
envisioned like a huge Pachinko hall, in which
balls are trickling down in parallel in hundreds or 
thousands of machines. It is highly 
likely that more than one solution is found
when there are a lot of trees, particularly 
if combinations of categories are allowed as well. 
So, an additional competitive process must 
take place to rank categories even 
though multiple solutions are offered to subsequent
verbalisation processes. We see therefore the same
characteristics as for the perceptual layer, and thus
the ``sieve architecture'' also 
applies for the conceptual layer (\figref{f:sieve}).

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (square) & 0.413 & 0.317  \\ 
1 (circle) & 0.410 & 0.410 \\ 
2 (square) & 0.163 & 0.163 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game8}Sensory data for the scene in \figref{scene2}.}
\end{center}
\end{table}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene8.pdf}}
\caption{\label{scene2}Top: The scene used in Game 8. 
Shape 0 is the topic. Bottom: The discrimination trees available
for this game.}
\end{figure}
The game is reported by the commentator as follows: 
\begin{verbatim}
Game 8
 a1 segments the context in 3 objects: 
    square-0, circle-1, square-2
 a1 chooses square-0 as the topic
 The discrimination game fails
\end{verbatim}
The game fails because for the two sensory channels 
for which there are discrimination trees, the values 
of the segments are all within the lower range and so 
no distinctive category or category set could be found. 
This failure stimulates
the discrimination network to expand, but any node, 
including a top node of some of the other sensory channels 
can be chosen for further expansion.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/scene22.pdf}}
\caption{\label{scene3}Top: The scene used in Game 22. 
Bottom: The discrimination trees available to the
agent.}
\end{figure}
The next example shows a game (Game 22) based on the scene 
in \figref{scene3}. The topic is the triangle, shape 0. 
The discrimination trees for 
HEIGHT and WIDTH have already more than one level and 
a discrimination tree for HPOS has developed.
The relevant sensor-scaled data for these three sensory
channels is shown in \tabref{tab:t-game22}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.167 & 0.437 & 0.573  \\ 
1 (rectangle) & 0.789 & 0.563 & 0.287 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}

The scene is very simple so there are several possible
solutions: The triangle is more to the left, it is less wide 
and taller. Each of these possibilities is discovered
and their score (purely based on past performance) is 
looked up. The game is reported by the commentator
as follows: 
\begin{verbatim}
Game 22
 a1 segments the scene in 2 objects:  
   triangle-0, rectangle-1
 a1 chooses triangle-0 as the topic 
 a1 categorises the topic as [HPOS 0.0–0.5] (score 0.57), 
   [HEIGHT 0.5–1.0] (score 0.09), or 
   [WIDTH 0.0–0.5] (score 0.0) 
 The discrimination game succeeds
\end{verbatim}

@@

Game 22 shows at once why context-scaling 
and saliency is important. When we inspect the scene in 
\figref{scene3}, we do not quite see so clearly 
that the triangle is less wide than the square, so why
is [WIDTH 0.0–0.5] 
nevertheless considered? Examination of the data shows
that the WIDTH values, 0.437 for the triangle
and 0.563 for the square, are very close 
to each other, but just by luck fall within 
the two regions carved out by the WIDTH discrimination
tree. On the other hand, the values on the HPOS channel 
are much further apart and so they are preferred.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
 0.622 & 0.125 & 0.286  \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22-sal}Sensory data for the scene in Game 22.}
\end{center}
\end{table}
From this table we see immediately that HPOS is the 
most salient channel and should be preferred by far, followed
by HEIGHT and then WIDTH. Thus we can expect the
agent to choose HPOS based on saliency. When the saliency threshold
is set to a reasonably high value, the other channels
would not even be considered, they would not pass the 
sieve of the perceptual layer.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HPOS} & {\itshape WIDTH} & {\itshape HEIGHT} \\ \midrule
0 (triangle) & 0.0 & 0.0 & 1.0  \\ 
1 (rectangle) & 1.0 & 1.0 & 0.0 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game22scaled}Sensory data for the scene shown in \figref{scene3}.}
\end{center}
\end{table}
So the best thing to do (and this is what the Talking Heads 
effectively do) is to first perform sensor-scaling, then 
compute saliency to determine which channel should be 
preferred, then perform context-scaling, to get clearly 
distinguished sensory values, and then do categorisation. 
Note that context-scaling has the same effect as using 
prototype-based categorisation because the actual values 
are pulled towards extremes, and thus perceived as 
prototypes. Context-scaling is not always desirable. For example, 
in the case of colour categorisation the actual channel
data should be maintained because here categorisation 
takes place on the basis of actual values.

@@

The next game (Game 24) is based on the scene in 
\figref{scene4}. The topic is triangle-0. 
The discrimination trees are the same as for
Game 22. The game (based on sensor-scaled values)
succeeds with a conjunctive combination of two
categories: 
\begin{verbatim}
Game 24
 a1 segments the scene in 4 objects: 
   triangle-0, triangle-1, square-2, rectangle-3 
 a1 chooses triangle-0 as topic
 a1 categorises the topic as 
   {[HEIGHT 0.0–0.5] [WIDTH 0.5–1.0]}
 The discrimination game succeeds
\end{verbatim}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap4/figs/game24.pdf}}
\caption{\label{scene4}The scene used in Game 24.}
\end{figure}
The relevant data after sensor-scaling for the 
two sensory channels involved are shown in \tabref{tab:t-game24}.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape Object} & {\itshape HEIGHT} & {\itshape WIDTH} \\ \midrule
0 (triangle) & 0.170 & 0.513  \\ 
1 (triangle) & 0.653 & 0.570 \\  
2 (square) & 0.213 & 0.213 \\  
3 (rectangle) & 0.613 & 0.310 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-game24}Sensory data for the scene in \figref{scene4}.}
\end{center}
\end{table}

[HEIGHT 0.0–0.5] is valid for triangle-0 and square-2 but
filters out the other segments. 
[WIDTH 0.5–1.0] is valid for triangle-0 and triangle-1 and 
filters out the others.  
The conjunctive combination of these two categories only 
retains triangle-0 and is therefore the one that is chosen.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & Saliency\\ \midrule
HPOS & 0.27 & 0.16 & 0.11\\ 
VPOS & 0.20 & 0.20 & 0.0\\ 
HEIGHT & 0.15 & 0.15 & 0.0\\ 
WIDTH & 0.10 & 0.11 & 0.01\\ 
AREA & 0.10 & 0.10 & 0.0\\ 
R & 0.23 & 0.25 & 0.02\\ 
G & 0.32 & 0.34  & 0.02\\ 
B & 0.63 & 0.65 & 0.02\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-real}Sensory data from a real world scene with segmentation shown in \ref{f:plate-10}.}
\end{center}
\end{table}
Clearly HPOS is the most salient channel and should be 
prefered by the agent. When performing context scaling, 
the two values for HPOS are drawn apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [HPOS 0.5–1.0] (to the 
right) easily distinguishes the topic (object-0) from
object-1. The game is reported as follows: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5–1.0]
\end{verbatim}
This example illustrates well why the categorisation 
of the Talking Heads is so robust and why agents 
often share the same conceptualisation even if the details  
of their raw perception is quite different. 
The saliency factor helps to focus the agents on those
aspects of the scene that stand out. There is 
an enormous reduction of variation, first by scaling then by 
the categorisation process itself.

@@

A new subcategoriser is constructed by taking a
categoriser node in the tree and dividing its range into 
two new subranges and thus two new subcategorisers.
For example, if there is a 
categoriser [HPOS 0.0–0.5], which triggers when the 
object is in the left most half of a scene, i.e. with HPOS
within $[0.0,0.5]$, then two 
subcategories are created by dividing $[0.0,0.5]$
into two halves, one for the 
range $[0.0,0.25]$ ([HPOS 0.0–0.25] or totally left) 
and one for the range $[0.25,0.5]$ ([HPOS 0.25--0.5] or mid-left). 
A new categoriser is added to the tree for each of these
halves.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap4/figs/avsuc.pdf}}
\caption{\label{avsuc}The graph displays the 
average success per 25 Discrimination Games for a
series of 500 games played by a single agent. Success climbs
to 100\%. The graph also displays the size of the 
agent's repertoire of categories. Each scene contains
between 3 and 6 objects.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.75\textwidth]{chap4/figs/dynatrees.pdf}}
\caption{\label{dyna-trees}Some snapshots of an evolving 
repertoire in a single agent using growth and pruning. 
Trees are shown after 500 games (a), 1000 games (b), 1500 games
(c) and 2000 games (d).} 
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/adptwrl1.pdf}}
\caption{\label{adptwrl1}Two snapshots 
from a series of 500 discrimination 
games developing categories to distinguish rectangles of 
the same average graylevel. Channel 2 (the ratio channel) 
and channel 3 (the grayscale channel) do not develop.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap4/figs/adptwrl2.pdf}}
\caption{\label{adptwrl2}Two snapshots from 
an additional series of 500 Discrimination Games after
the environment has become more complex. RATIO
(channel 2) and GRAY (channel 3) have started to develop.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap4/figs/adptwrlglo.pdf}}
\caption{\label{adptwrlglo}Evolution of success and ontology
size in a series of 1000 Discrimination Games played by a single agent.
In phase 1, only rectangles of the same graylevel are 
generated by the GEOM world. In phase 2, additional types of 
shapes are generated by the environment.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap4/figs/discri200.pdf}}
\caption{\label{discri200a}The discrimination trees 
developed by two physically embodied agents 
{\bfshape a1} (left) and {\bfshape a2} (right). The top of the 
figure shows the trees after 
playing 100 games and bottom after 200 games.}
\end{figure}
